---
title: "6: Part 1 - Generalized Linear Models"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Fall 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Describe the components of the generalized linear model (GLM)
2. Apply special cases of the GLM (linear regression) to real datasets
3. Interpret and report the results of linear regressions in publication-style formats
3. Apply model selection methods to choose model formulations

## Generalized Linear Models (GLMs)

The analysis of variance (ANOVA), and linear regression are special cases of the **generalized linear model** (GLM). The GLM also includes analyses not covered in this class, including logistic regression, multinomial regression, chi square, and log-linear models. The common characteristic of general linear models is the expression of a continuous response variable as a linear combination of the effects of categorical or continuous explanatory variables, plus an error term that expresses the random error associated with the coefficients of all explanatory variables. The explanatory variables comprise the deterministic component of the model, and the error term comprises the stochastic component of the model. Historically, artificial distinctions were made between linear models that contained categorical and continuous explanatory variables, but this distinction is no longer made. The inclusion of these models within the umbrella of the GLM allows models to fit the main effects of both categorical and continuous explanatory variables as well as their interactions. 

### Choosing a model from your data: A "cheat sheet"

**One-way ANOVA (Analysis of Variance):** Continuous response, one categorical explanatory variable with more than two categories.
  When the categorical explanatory variable just have two categories use the t-test

**Two-way ANOVA (Analysis of Variance)** Continuous response, two categorical explanatory variables.

**Single Linear Regression** Continuous response, one continuous explanatory variable.

**Multiple Linear Regression** Continuous response, two or more continuous explanatory variables.

**ANCOVA (Analysis of Covariance)** Continuous response, categorical explanatory variable(s) and  continuous explanatory variable(s).

If multiple explanatory variables are chosen, they may be analyzed with respect to their **main effects** on the model (i.e., their separate impacts on the variance explained) or with respect to their **interaction effects,** the effect of interacting explanatory variables on the model. 

### Hypothesis testing
Make an assumption about a population parameter that may or may not be true. Then test the null hypothesis and the alternative hypothesis. 

Procedure: state the hypotheses and identify the claim. Then, find the critical values, compute the test value, and make the decision to reject or not reject the null hypothesis. If the p value is less than the alpha level, reject the null hypothesis. 

### Linear regression
Trying to compare a variable Y with a variable X, trying to understand the relationship between them. Find the equation that minimizes the sum of squared vertical deviations from the line. X axis is the explanatory variable. Explain Y with X. Linear regression allows you to build the blue straight line; equation for that line that explains. 

### Standard error
Measure of the distance of the points around that regression line. Want the standard error to be as small as possible. Scatter of points are deviating around the line, but the deviations are small. 

### Correlation coefficient
Measure of the strength and the direction of the relationship between two variables. Positive correlation (slope is positive). Negative correlation means if one variable increases, the other will decrease. r^2 is a measure of the percentage of variability in the values of Y that is explained by X. They range from 0-1. R^2 close to 1 means almost all variability of Y is explained by X. 

### Residuals
After fitting the model, we need to check the residuals. Do our residuals follow the normal distribution? Do we have unbiased estimates?

### P-values and coefficinets
P-value for coefficients will indicate if the relationship is statistically significant. P value for each independent value tests the null hypothesis that there is no correlation between the variables. If the p-value is less, then you can reject the null hypothesis for the entire population. 

### Assumptions of the GLM

The GLM is based on the assumption that the data residuals approximate a normal distribution (or a linearly transformed normal distribution). We will discuss the non-parametric analogues to several of these tests if the assumptions of normality are violated. For tests that analyze categorical explanatory variables, the assumption is that the variance in the response variable is equal among groups. Note: environmental data often violate the assumptions of normality and equal variance, and we will often proceed with a GLM even if these assumptions are violated. In this situation, justifying the decision to proceed with a linear model must be made.

## Set up
```{r, message = FALSE}
getwd()
library(tidyverse)
options(scipen = 4)

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Linear Regression
A linear regression is comprised of a continuous response variable, plus a combination of 1+ continuous response variables (plus the error term). The deterministic portion of the equation describes the response variable as lying on a straight line, with an intercept and a slope term. The equation is thus a typical algebraic expression: 
$$ y = \alpha + \beta*x + \epsilon $$
Alpha = y intercept, beta = slope
Always have error = residual error term, will always have points away from the line
The goal for the linear regression is to find a **line of best fit**, which is the line drawn through the bivariate space that minimizes the total distance of points from the line. This is also called a "least squares" regression. The remainder of the variance not explained by the model is called the **residual error.** 

The linear regression will test the null hypotheses that

1. The intercept (alpha) is equal to zero.
2. The slope (beta) is equal to zero

Whether or not we care about the result of each of these tested hypotheses will depend on our research question. Sometimes, the test for the intercept will be of interest, and sometimes it will not.

Important components of the linear regression are the correlation and the R-squared value. The **correlation** is a number between -1 and 1, describing the relationship between the variables. Correlations close to -1 represent strong negative correlations, correlations close to zero represent weak correlations, and correlations close to 1 represent strong positive correlations. The **R-squared value** is the correlation squared, becoming a number between 0 and 1. The R-squared value describes the percent of variance accounted for by the explanatory variables. 

## Simple Linear Regression

For the NTL-LTER dataset, can we predict irradiance (light level) from depth?

```{r}
irradiance.regression <- lm(PeterPaul.chem.nutrients$irradianceWater ~ PeterPaul.chem.nutrients$depth)
#lm stands for linear model

# another way to format the lm() function, preferred way
irradiance.regression <- lm(data = PeterPaul.chem.nutrients, irradianceWater ~ depth)
summary(irradiance.regression)
#slope of the line is -95.89, means there is a negative relationship. 
#p-value is less than 0.05
#r-square: depth explains around 30% of the variance in irradiance

# Correlation
cor.test(PeterPaul.chem.nutrients$irradianceWater, PeterPaul.chem.nutrients$depth)
#negative correlation, -0.55 strong correlation
#lower light levels at greater depths
```
Question: How would you report the results of this test (overall findings and report of statistical output)?

>  

So, we see there is a significant negative correlation between irradiance and depth (lower light levels at greater depths), and that this model explains about 31 % of the total variance in irradiance. Let's visualize this relationship and the model itself. 

An exploratory option to visualize the model fit is to use the function `plot`. This function will return four graphs, which are intended only for checking the fit of the model and not for communicating results. The plots that are returned are: 

1. **Residuals vs. Fitted.** The value predicted by the line of best fit is the fitted value, and the residual is the distance of that actual value from the predicted value. By definition, there will be a balance of positive and negative residuals. Watch for drastic asymmetry from side to side or a marked departure from zero for the red line - these are signs of a poor model fit.
#good fit means there's a balance in residuals (pos and neg)

2. **Normal Q-Q.** The points should fall close to the 1:1 line. We often see departures from 1:1 at the high and low ends of the dataset, which could be outliers. 
#compare the residuals to a normal distribution. Points should follow closesly. 

3. **Scale-Location.** Similar to the residuals vs. fitted graph, this will graph the squared standardized residuals by the fitted values. 
#similar to residual vs. fitted

4. **Residuals vs. Leverage.** This graph will display potential outliers. The values that fall outside the dashed red lines (Cook's distance) are outliers for the model. Watch for drastic departures of the solid red line from horizontal - this is a sign of a poor model fit.

```{r, fig.height = 3, fig.width = 4}
par(mfrow = c(2,2), mar=c(4,4,4,4)) #dividing the plot screen into four cells, transforming screen into a grid. 
plot(irradiance.regression)
par(mfrow = c(1,1))
```

The option best suited for communicating findings is to plot the explanatory and response variables as a scatterplot. 

```{r, fig.height = 3, fig.width = 4}
# Plot the regression
irradiancebydepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
  ylim(0, 2000) +
  geom_point() 
print(irradiancebydepth) 
```

Given the distribution of irradiance values, we don't have a linear relationship between x and y in this case. Let's try log-transforming the irradiance values. In other words, we take the log of the irradiance observations and compare to our depth levels. Note we also removing the observations that seems to ba an outlier.

```{r, fig.height = 3, fig.width = 4}
PeterPaul.chem.nutrients <- filter(PeterPaul.chem.nutrients, 
                                   irradianceWater != 0 & irradianceWater < 5000) #to take the log we need to eliminate zeros and get rid of outliers
irradiance.regression2 <- lm(data = PeterPaul.chem.nutrients, log(irradianceWater) ~ depth)
summary(irradiance.regression2) #notice that the relationship is still negative
#notice that the residual standard error is reduced and the R-squared is now explaining 73 percent of the data. Much better fit than what we had before. 

par(mfrow = c(2,2), mar=c(4,4,4,4))
plot(irradiance.regression2)
par(mfrow = c(1,1))
#Now we can see symmetry a bit better. 

# Add a line and standard error for the linear regression
irradiancebydepth2 <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
  geom_smooth(method = "lm") +
  scale_y_log10() +
  geom_point() 
print(irradiancebydepth2) 
#scale_y_log10 allows us to take the log

# SE - confidence interval around smooth can also be removed
irradiancebydepth2 <- 
    ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
    geom_point() +
    scale_y_log10() +  #log transformed values for irradianceWater
    geom_smooth(method = 'lm', se = FALSE, color = "black")
print(irradiancebydepth2)

# Make the graph attractive

```

## Multiple Linear Regression
It is possible, and often useful, to consider multiple continuous explanatory variables at a time in a linear regression. For example, total phosphorus concentration in Paul Lake (the unfertilized lake) could be dependent on depth and dissolved oxygen concentration: #More than one x explaining variability in Y

``` {r, fig.height = 3, fig.width = 4}
TPregression <- lm(data = subset(PeterPaul.chem.nutrients, lakename == "Paul Lake"), 
                   tp_ug ~ depth + dissolvedOxygen)
summary(TPregression)
#All values are less than 0.05, so the relationship is different from zero among the two x and the y individually. We can also see that our r^2 is 0.29, only 30% is due to the depth and the dissolved oxygen concentration. 


TPplot <- ggplot(subset(PeterPaul.chem.nutrients, lakename == "Paul Lake"), 
                 aes(x = dissolvedOxygen, y = tp_ug, color = depth)) +
  geom_point() +
  xlim(0, 20)
print(TPplot)

par(mfrow = c(2,2), mar=c(4,4,4,4))
plot(TPregression)
par(mfrow = c(1,1))
#first - look for symmetry with respect to the red line, and want it to be flat. But we are seeing not a good residual series, something is missing. Maybe we need more variables. 
#second - not too bad, some deviations on high and low
#third - same issues with residuals
#fourth - some of the observations here seem to be outliers

```

## Correlation Plots
We can also make exploratory plots of several continuous data points to determine possible relationships, as well as covariance among explanatory variables. # Want to look for more explanatory variables. # One way to find relevant variables is by looking at correlation coefficients. We have so many variables that we may want to create a correlation matrix. We can see the correlation among all the data we have. 

```{r, fig.height = 3, fig.width = 4}
#install.packages("corrplot")
library(corrplot)
PeterPaulnutrients <- 
  PeterPaul.chem.nutrients %>%
  select(tn_ug:po4) %>%
  na.omit()
PeterPaulCorr <- cor(PeterPaulnutrients) #correlation matrix
corrplot(PeterPaulCorr, method = "ellipse") #thinner means higher correlation, color is changing from red (-1) to (+1)
corrplot.mixed(PeterPaulCorr, upper = "ellipse")
```

## AIC to select variables

However, it is possible to over-parameterize a linear model. Adding additional explanatory variables takes away degrees of freedom, and if explanatory variables co-vary the interpretation can become overly complicated. Remember, an ideal statistical model balances simplicity and explanatory power! To help with this tradeoff, we can use the **Akaike's Information Criterion (AIC)** to compute a stepwise regression that either adds explanatory variables from the bottom up or removes explanatory variables from a full set of suggested options. The smaller the AIC value, the better. 

Let's say we want to know which explanatory variables will allow us to best predict total phosphorus concentrations. Potential explanatory variables from the dataset could include depth, dissolved oxygen, temperature, PAR, total N concentration, and phosphate concentration.

```{r}
Paul.naomit <- PeterPaul.chem.nutrients %>%
  filter(lakename == "Paul Lake") %>%
  na.omit()

TPAIC <- lm(data = Paul.naomit, tp_ug ~ depth + dissolvedOxygen + 
              temperature_C + tn_ug + po4) #considering all 5

#Choose a model by AIC in a Stepwise Algorithm
step(TPAIC) #best regression is tn_ug, oxygen, and temperature
TPmodel <- lm(data = Paul.naomit, tp_ug ~ dissolvedOxygen + temperature_C + tn_ug)
summary(TPmodel)

```