---
title: "7: Time Series"
author: "Environmental Data Analytics | Adapted by John Fay and Luana Lima | Developed by Kateri Salk"
date: "Fall 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Discuss the purpose and application of time series analysis for environmental data
2. Explore the components of times series: trend, seasonal, random
3. Learn how to perform stationarity test

## Notes on Lecture
1. Time series are a set of observations on a variable that were collected over time. Could be discrete or continuous. Stock prices, retail sales. We have mathematical respresentations; defined by values Y. Index variables as Y1, Y2, and so forth. At time 1, time 2. Y will be a function of Y. Time series analysis (TSA), analyze the past behavior to predict future behavior. Causes of variation include seasons, holidays, natural calamities. Looking at the plot, you can start to make some observations. Hard to make conclusions. See a trend in which the black line is decreasing as the red line is increasing. Emissions are going to be decreasing as we use renewables. Project percentage growth for renewables in the future after looking at the past relationships between renewables and emissions. Electricity consumption for a specific smart meter. You can see the pattern in the TSA, in which consumption goes up and down. Difference between night and day (day our consumption peaks). Helpful for determining forecasted activities/planning for future operations. Money we spent generating electricity could have been saved if we knew future scenarios. Make better decisions. Time frames: short, medium, and long-term horizon. How far can we predict? Another component is trend: tendency to grow or decline. Easy to detect from plot. Cycles are up and down repetitive moments. Seasonal variation (don't mix with cycle - this is up and down movement but occurs more periodically - short duration). Clear seasonal variation. Random variations are difficult; eratic movements that are not predictable because they don't follow a pattern. 

Stationary data: no significant upward or downward trend.

Nonstationary data: time series that will have a significant upward or downward trend over time.

Seasonal: regular patterns over time.

What is correlation? Correlation is a measure of linear dependence between two variables. In TSA: autocorrelation is the measure of dependence between two adjacent values of the same variables. Self correlation. Y(t) is the original series and Y(s) is a lagged version of the series. Comparing observations in regular Y with previous observation and seeing how they relate. Lag 1 autocorrelation is the correlation between Y(t) and Y(s). Lag 2, then, compare Y(1) with Y(3) instead of Y(2). Autocovariance and autocorrelation function give information about the dependence structure of a time series. Good model for time series will need to have the observation at T-1 and T-2. What happes two steps before T will be relevant to T. 

Stationary process: laws that govern the behavior of the process do not change over time. No matter where in time, same mean, SD, and autocorrelation. White noise is a great example of a stationary process. It is a sequence of independent, randomly distributed variables. Mean =0. 

Partial autocorrelation: ACF (autocorrelation function) --> measures the linear dependency of Y(t) and Y(t-h). Intermediate variables also play an important value. 

Partial Autocorrelation: Remove the influence of all of these intermediate variables --> only a direct correlation. 

You always build the ACF and PACF to a stochastic process. Autoregressive: regression on previous observations. No external variables, only lagged variables. 

PACF is the correlation between lagged variables when we remove the intermediate. ACF and PACF for lag 1 will always be the same, but lag 2 will be different. 

ACF --> seasonal dataset.

Trend component: long term tendency that you see on your dataset. Could be linear or nonlinear. 

Time series plot: able to identify if there is a trend or not (example with clear upward linear trend). Increasing trend. We try to start modeling the data set with a linear trend. If that does not work well, we can try to improve that. 

Linear trend: Y=mx+b+error
slope, intercept, and error. 

Error: distance from observation to trend. 

We estimate the trend because we want to remove the trend; want the mean to be the same. 

Removal: Y - estimated trend. This is your error. We are creating a series of random trends. 

Non-linear trends: if you have a quadratic trend, you can fit the curve to the data, model the distances. Exponential trends can be represented by linear trends by taking the log of your series. 

Seasonal components: Short term regular wavelength pattern. Trend over time of increase on the total demand. Seasonal component going up and down and a trend over time. Overall, the demand increases.

Estimating seasonal trend: seasonal means model. Series represented as a sum of U(t) + x. Monthly seasonal data (have 12 parameters). U(t) will have 12 values. Beta 1-12. How can I remove the seasonal trend? Take our y(t) and subtract the seasonal component we estimated. 

Stochastic vs. deterministic trend: Detrending is accomplished by running a regression and obtaining the residuals. And the residuals will be the detrended series. Trend-stationary. 

You can have a stochastic trend: has different behaviors over time. During specific times, the slope of the trend was different. A trend that changes over time. Cannot be removed by the same process as deterministic trend.

Data set with trend in blue. Represented as dashed black line. Stochastic is represented by green line, harder to fit a linear trend. 

Mann-Kendall test: detect deterministic †rends in a series of environmental data, climate data, or hydrological data. Cannot be applied to seasonal data. 

H0: Y(t) is stationary
H1: Y(t) follows a trend

Test checks the magnitude of S.

Spearman rank: does not have to be linear. Compute between Y(t) and t and if there is a strong correlation there is a trend. Cor or Cor.test will generate the coefficient. 

Dick-fuller test: applied to verify the stochastic trend or to verify if the time series has a unit root. Dealing with a stochastic trend, dealing with a series that has a unit root. What is a unit root? The objective is to test if 
H0: contains a unit root
H1: is stationary (will eventually go back to mean a). Whenever you have a unit root, it will be hard to control. Differentiate. 

ADF.test --> comments about this test.

## Set up

Today we will work with two datasets. The USGS dataset on discharge at the Eno River and a new dataset we haven't explored yet on wind speed. The data file is available at "./Data/Raw/Wind_Speed_PortArthurTX.csv". It contains average wind speed in monthly time steps (elevation = 5 meters). The data is available from NOAA National Centers for Environmental Information (NCEI) [here][https://www.ncdc.noaa.gov/cdo-web/datasets#GSOM].

```{r, message = FALSE}

library(tidyverse)
library(ggplot2)
library(lubridate)
install.packages("trend")
library(trend)
install.packages("zoo")
library(zoo)
#install.packages("Kendall")
library(Kendall)
#install.packages("tseries")
library(tseries)

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

#Read Eno river data
EnoDischarge <- read.csv("./Data/Processed/USGS_Site02085000_Flow_Processed.csv",
                         stringsAsFactors = TRUE)
EnoDischarge$datetime <- as.Date(EnoDischarge$datetime, format = "%Y-%m-%d")

#Read wind speed data
wind_data <- read.csv(file="./Data/Raw/Wind_Speed_PortArthurTX.csv",header=TRUE,
                      stringsAsFactors = TRUE)
wind_data$DATE <- ym(wind_data$DATE)

```


## Time Series Analysis overview

Time series are a special class of dataset, where a response variable is tracked over time. The frequency of measurement and the timespan of the dataset can vary widely. At its most simple, a time series model includes an explanatory time component and a response variable. Mixed models can include additional explanatory variables (check out the `nlme` and `lme4` R packages). We will cover a few simple applications of time series analysis in these lessons, with references for how to take analyses further.

### Opportunities

Analysis of time series presents several opportunities. For environmental data, some of the most common questions we can answer with time series modeling are:

* Has there been an increasing or decreasing **trend** in the response variable over time?
* Can we **forecast** conditions in the future?

### Challenges

Time series datasets come with several caveats, which need to be addressed in order to effectively model the system. A few common challenges that arise (and can occur together within a single dataset) are: 

* **Autocorrelation**: Data points are not independent from one another (i.e., the measurement at a given time point is dependent on previous time point(s))
* **Data gaps**: Data are not collected at regular intervals, necessitating *interpolation* between measurements.
* **Seasonality**: seasonal patterns in variables occur at regular intervals, impeding clear interpretation of a monotonic (unidirectional) trend.
* **Heteroscedasticity**: The variance of the time series is not constant over time
* **Covariance**: the covariance of the time series is not constant over time

## Handling data gaps and missing data. Example: Eno River Discharge

River discharge is measured daily at the Eno River gage station. Since we are working with one location measured over time, this will make a great example dataset for time series analysis. 

Let's look at what the dataset contains for mean daily discharge.

```{r}
ggplot(EnoDischarge, aes(x = datetime, y = discharge.mean)) +
  geom_line() +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)"))
```

Notice there are missing data from 1971 to 1985. Gaps this large are generally an issue for time series analysis, as we don't have a continuous record of data or a good way to characterize any variability that happened over those years. We will illustrate a few workarounds to address these issues. 

Let's start by removing the NAs and splitting the dataset into the early and late years. 

```{r}
EnoDischarge.complete <- EnoDischarge %>%
  drop_na(discharge.mean)

EnoDischarge.early <- EnoDischarge.complete %>%
  filter(datetime < as.Date("1985-01-01")) #less than 1985

EnoDischarge.late <- EnoDischarge.complete %>%
  filter(datetime > as.Date("1985-01-01")) #greater than 1985
```

## Decomposing a time series dataset

A given time series can be made up of several component series: 

1. A **seasonal** component, which repeats over a fixed known period (e.g., seasons of the year, months, days of the week, hour of the day)
2. A **trend** component, which quantifies the upward or downward progression over time. The trend component of a time series does not have to be monotonic.
3. An **error** or **random** component, which makes up the remainder of the time series after other components have been accounted for. This component reflects the noise in the dataset. 
4. (optional) A **cyclical** component, which repeats over periods greater than the seasonal component. A good example of this is El Niño Southern Oscillation (ENSO) cycles, which occur over a period of 2-8 years.

### Example: Eno discharge

We will decompose the EnoDischarge.late data frame for illustrative purposes today. It is possible to run time series analysis on detrended data by subtracting the trend component from the data. However, detrending must be done carefully, as many environmental data are bounded by zero but are not treated as such in a decomposition. If you plan to use decomposition to detrend your data, please consult time series analysis guides before proceeding.

We first need to turn the discharge data into a time series object in R. This is done using the `ts` function. Notice we can only specify one column of data and need to specify the period at which the data are sampled. The resulting time series object cannot be viewed like a regular data frame.

Note: time series objects must be equispaced. In our case, we have daily data with no NAs in the data frame, so we don't need to worry about this. We will cover how to address data that are not equispaced later in the lesson.

```{r}
EnoDischarge.late_ts <- ts(EnoDischarge.late$discharge.mean, start = c(1985,1), frequency = 365) #defining starting point (first observation, 1985 for us -- put in year,month frequency = 365.)
```

The `stl` function decomposes the time series object into its component parts. We must specify that the window for seasonal extraction is either "periodic" or a specific number of at least 7. The decomposition proceeds through a loess (locally estimated scatterplot smoothing) function.

```{r}
?stl
# Generate the decomposition
EnoDischarge.late_Decomposed <- stl(EnoDischarge.late_ts, s.window = "periodic") #only 2 variables

# Visualize the decomposed series. 
plot(EnoDischarge.late_Decomposed)
#first observed, #second seasonal, #third trend, #fourth remainder.Seasonal repeats itself every 365 observations. For trend, some periods where increased, some decreased. Definitely not a monotonic trend.

# We can extract the components and turn them into data frames
EnoDischarge.late_Components <- as.data.frame(EnoDischarge.late_Decomposed$time.series[,1:3])

EnoDischarge.late_Components <- mutate(EnoDischarge.late_Components,
        Observed = EnoDischarge.late$discharge.mean,     
        Date = EnoDischarge.late$datetime)
#adding two more columns with observed data and date objects

# Visualize how the trend maps onto the data
ggplot(EnoDischarge.late_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Discharge (ft"^3*"/s)"))

# Visualize how the seasonal cycle maps onto the data
ggplot(EnoDischarge.late_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Discharge (ft"^3*"/s)")) #going below y=0 line in every cycle of the seasonal component

```

Note that the decomposition can yield negative values when we apply a seasonal adjustment or a trend adjustment to the data. The decomposition is not constrained by a lower bound of zero as discharge is in real life. Make sure to interpret with caution!


## Trend analysis

Two types of trends may be present in our time series dataset: **monotonic/deterministic** or **stochastic**. Monotonic trends are a gradual shift over time that is consistent in direction, for example in response to land use change. 

A third type of trend we haven't talked about is the **step** trend, also knows as a level shift. Step trends are a distinct shift at a given time point, for example in response to a policy being enacted. 


### Monotonic trend analysis

In general, detecting a monotonic trend requires a long sequence of data with few gaps. If we are working with monthly data, a time series of at least five years is recommended. Gaps can be accounted for, but a gap that makes up more than 1/3 of the sampling period is generally considered the threshold for considering a gap to be too long (a step trend analysis might be better in this situation). 

Adjusting the data may be necessary to fulfill the assumptions of a trend test. A common method to replace missing values is **interpolation**. Common interpolation methods: 

* **Piecewise constant**: also known as a "nearest neighbor" approach. Any missing data are assumed to be equal to the measurement made nearest to that date (could be earlier or later).
* **Linear**: could be thought of as a "connect the dots" approach. Any missing data are assumed to fall between the previous and next measurement, with a straight line drawn between the known points determining the values of the interpolated data on any given date.
* **Spline**: similar to a linear interpolation except that a quadratic function is used to interpolate rather than drawing a straight line.

#important to know if there is any missing data on your dataset. If there is a big gap, that challenges 
#if the gap is not too long, we can use interpolation.
#piecewise constant (nearest neighbor, any missing data will be assumed to be equal to nearest measurement, earlier or later)
#linear interpolation (like connecting the dots, and then there's a gap, the linear interpolation will connect the dots). Any missing data will fall between (EMPHASIS ON BETWEEN) nearest and next.
#spline uses a quadratic function rather than a straight line.

#### Example: interpolation

The Eno River discharge data doesn't have any short periods of missing data, so interpolation would not be a good choice for that dataset. We will illustrate a linear interpolation using the wind speed dataset. 


```{r}
head(wind_data)
summary(wind_data$AWND)

# Adding new column with no missing obs, just for illustration purpose
# In real applications you will simply replace NAs
wind_data_clean <- 
  wind_data %>% 
  mutate( AWND.clean = zoo::na.approx(AWND) ) #connected dots
#use NA approx if the data is in the middle of the data set (not at end or beginning)

summary(wind_data_clean$AWND.clean)
#Note the NA is gone

ggplot(wind_data_clean ) +
  geom_line(aes(x = DATE, y = AWND.clean), color = "red") +
  geom_line(aes(x = DATE, y = AWND), color = "black") +
  ylab("Average Wind Speed") 
```

### Monotonic trend analysis, continued

Specific tests for monotonic trend analysis are listed below, with assumptions and tips: 

* **linear regression**: no seasonality, fits the assumptions of a parametric test. Function: `lm`
* **Mann-Kendall**: no seasonality, non-parametric, missing data allowed. Function: `MannKendall()` (package: Kendall)
* **Seasonal Mann-Kendall**: seasonality, non-parametric `SeasonalMannKendall` (package: Kendall)
* **Spearman Rho**: no seasonality, non-parametric, missing data allowed. Function: `cor.test(method="spearman")` (package: stats)

Specific test for stochastic trend analysis:

* **Augmented Dickey Fuller**: no seasonality, non-parametric, missing data not allowed. Function: `adf.test()` (package: tseries)


#### Example: monotonic trend analysis

Let's refer to our wind speed data. We already performed interpolation, but we still need to create our time series object and decompose the series to find out which stationarity test we can apply. 

Note that wind speed has a seasonal cycle. We might be interested in knowing how (if) speed has changed over the course of measurement while incorporating the seasonal component. In this case, we will use a Seasonal Mann-Kendall test to figure out whether a monotonic trend exists.


```{r}

# Generate time series (trend test needs ts, not data.frame)
f_month <- month(first(wind_data_clean$DATE))
f_year <- year(first(wind_data_clean$DATE))
wind_data_ts <- ts(wind_data_clean$AWND.clean,
                   start=c(f_year,f_month),
                   frequency=12) 

#decompose
wind_data_decomp <- stl(wind_data_ts,s.window = "periodic")
plot(wind_data_decomp) #seasonal component is easy to spot

# Run SMK test
wind_data_trend1 <- Kendall::SeasonalMannKendall(wind_data_ts)

# Inspect results
wind_data_trend1
summary(wind_data_trend1) #reject the null hypothesis that data is stationary --> reject and state that we have a trend


wind_data_trend2 <- trend::smk.test(wind_data_ts)
# Inspect results
wind_data_trend2
summary(wind_data_trend2) #can see individual seasons

#Visualization
wind_data_plot <-
ggplot(wind_data, aes(x = DATE, y = AWND)) +
  geom_point() +
  geom_line() +
  ylab("Average Wind Speed") +
  geom_smooth( method = lm )
print(wind_data_plot) #decreased trend over time

```

What would we conclude based on these findings? 

> Answer: Monotonic trend exists! Decreases.


## Forecasting with Autoregressive and Moving Average Models (ARMA)

We might be interested in characterizing a time series in order to understand what happened in the past and to effectively forecast into the future. Two common models that can approximate time series are **autoregressive** and **moving average** models. To classify these models, we use the  **ACF (autocorrelation function)** and the **PACF (partial autocorrelation function)**, which correspond to the autocorrelation of a series and the correlation of the residuals, respectively. 

**Autoregressive** models operate under the framework that a given measurements is correlated with  previous measurements. For example, an AR1 formulation dictates that a measurement is dependent on the previous measurement, and the value can be predicted by quantifying the lag. 

**Moving average** models operate under the framework that the covariance between a measurement and the previous measurement is zero. While AR models use past forecast *values* to predict future values, MA models use past forecast *errors* to predict future values.

Here are some great resources for examining ACF and PACF lags under different formulations of AR and MA models. 
https://nwfsc-timeseries.github.io/atsa-labs/sec-tslab-autoregressive-ar-models.html
https://nwfsc-timeseries.github.io/atsa-labs/sec-tslab-moving-average-ma-models.html

ARMA models require stationary data. This means that there is no monotonic trend over time and there is also equal variance and covariance across the time series. The function `adf.test` will determine whether our data are stationary. The null hypothesis is that the data are not stationary, so we infer that the data are stationary if the p-value is < 0.05.

While some processes might be easy to identify, it is often complicated to predict the order of AR and MA processes. To get around this issue, it is often necessary to run multiple potential formulations of the model and see which one results in the most parsimonious fit using AIC. The function `auto.arima` does this automatically.

