---
title: "4: Part 1 - Data Wrangling"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Fall 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Describe the usefulness of data wrangling and its place in the data pipeline
2. Wrangle datasets with dplyr functions
3. Apply data wrangling skills to a real-world example dataset

## Set up your session

Today we will work with a dataset from the [North Temperate Lakes Long-Term Ecological Research Station](https://lter.limnology.wisc.edu/about/overview). The NTL-LTER is located in the boreal zone in northern Wisconsin, USA. We will use the [chemical and physical limnology dataset](https://lter.limnology.wisc.edu/content/cascade-project-north-temperate-lakes-lter-core-data-physical-and-chemical-limnology-1984), running from 1984-2016. 

Opening discussion: why might we be interested in long-term observations of temperature, oxygen, and light in lakes?

> Add notes here: Climate change causes drying in lakes, lake drying is caused by increased temperature which evaporates the lakes. This changes the nutrient densities. Temperature, oxygen, and light are all abiotic factors that are incredibly important for biological species (lake vegetation, fish).

```{r, message = FALSE}
getwd()
#install.packages(tidyverse)
library(tidyverse)
#install.packages(lubridate)
library(lubridate)
NTL.phys.data <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

colnames(NTL.phys.data)
head(NTL.phys.data)
summary(NTL.phys.data)
str(NTL.phys.data)
dim(NTL.phys.data)

class(NTL.phys.data$sampledate)
# Format sampledate as date
NTL.phys.data$sampledate <- as.Date(NTL.phys.data$sampledate, format = "%m/%d/%y")
```

## Data Wrangling

Data wrangling extends data exploration: it allows you to process data in ways that are useful for you. An important part of data wrangling is creating *tidy datasets*, with the following rules: 

1. Each variable has its own column
2. Each observation has its own row
3. Each value has its own cell

What is the best way to wrangle data? There are multiple ways to arrive at a specific outcome in R, and we will illustrate some of those approaches. Your goal should be to write the simplest code that will get you to your desired outcome. However, there is sometimes a trade-off of the opportunity cost to learn a new formulation of code and the time it takes to write complex code that you already know. Remember that the best code is one that is easy to understand for yourself and your collaborators. Remember to comment your code, use informative names for variables and functions, and use reproducible methods to arrive at your output.

## Dplyr Wrangling Functions

`dplyr` is a package in R that includes functions for data manipulation (i.e., data wrangling or data munging). `dplyr` is included in the tidyverse package, so you should already have it installed on your machine. The functions act as verbs for data wrangling processes. For more information, run this line of code:

```{r, results = "hide"}
vignette("dplyr")
```

### Filter

Filtering allows us to choose certain rows (observations) in our dataset.

Here are the relevant commands used in the `filter` function. Add some notes to designate what these commands mean. 
`==` compare your value to another value
`!=` are they different?
`<` less than 
`<=` less than or equal to
`>` greater than
`>=` greater than or equal to
`&` multiple to check, both true
`|` multuple to check, one or the other should be true

```{r}
class(NTL.phys.data$lakeid) #mine says factor here
class(NTL.phys.data$depth)

# matrix filtering
NTL.phys.data.surface1 <- NTL.phys.data[NTL.phys.data$depth == 0,] #filtering rows with zero column depth

# dplyr filtering (to do the same thing)
NTL.phys.data.surface2 <- filter(NTL.phys.data, depth == 0)
NTL.phys.data.surface3 <- filter(NTL.phys.data, depth < 0.25) #allows you to include a filtered range

# Did the methods arrive at the same result?
head(NTL.phys.data.surface1)
dim(NTL.phys.data.surface1)
head(NTL.phys.data.surface2)
dim(NTL.phys.data.surface2)
head(NTL.phys.data.surface3)
dim(NTL.phys.data.surface3)
#yes they are the same

# Choose multiple conditions to filter
summary(NTL.phys.data$lakename) #number of occurrences
NTL.phys.data.PeterPaul1 <- filter(NTL.phys.data, lakename == "Paul Lake" | lakename == "Peter Lake") #select all instances of peter lake or paul lake. 
NTL.phys.data.PeterPaul2 <- filter(NTL.phys.data, lakename != "Central Long Lake" & 
                                     lakename != "Crampton Lake" & lakename != "East Long Lake" &
                                     lakename != "Hummingbird Lake" & lakename != "Tuesday Lake" &
                                     lakename != "Ward Lake" & lakename != "West Long Lake")
#opposite strategy, check for when they are not any of the other lakes, use !=
NTL.phys.data.PeterPaul3 <- filter(NTL.phys.data, lakename %in% c("Paul Lake", "Peter Lake"))
#percentage in percentage, useful when you need to select based on more than two elements. lead to same results and cleaner.

# Choose a range of conditions of a numeric or integer variable
summary(NTL.phys.data$daynum)
NTL.phys.data.JunethruOctober1 <- filter(NTL.phys.data, daynum > 151 & daynum < 305) #combination of both, why we use &
NTL.phys.data.JunethruOctober2 <- filter(NTL.phys.data, daynum > 151, daynum < 305) #comma means combination of both
NTL.phys.data.JunethruOctober3 <- filter(NTL.phys.data, daynum >= 152 & daynum <= 304) #make sure you adjust if you're using the equal sign
NTL.phys.data.JunethruOctober4 <- filter(NTL.phys.data, daynum %in% c(152:304)) #also can use %in%, thing about a value matching operator

#first day of June is 152, Nov 1 is 305

# Exercise for synchronus section: 
# filter NTL.phys.data for the year 1999
# what code do you need to use, based on the class of the variable?
class(NTL.phys.data$year4)

# Exercise for synchronus section: 
# filter NTL.phys.data for Tuesday Lake from 1990 through 1999.


```
Question: Why don't we filter using row numbers?

> Answer: 

### Arrange

Arranging allows us to change the order of rows in our dataset. By default, the arrange function will arrange rows in ascending order.

```{r}
NTL.phys.data.depth.ascending <- arrange(NTL.phys.data, depth) #specify the column we want to arrange by, here we are saying depth and ascending
NTL.phys.data.depth.descending <- arrange(NTL.phys.data, desc(depth)) #desc allows you to do descending

# Exercise: 
# Arrange NTL.phys.data by temperature, in descending order. 
# Which dates, lakes, and depths have the highest temperatures?


```
### Select

Selecting allows us to choose certain columns (variables) in our dataset.

```{r}
NTL.phys.data.temps <- select(NTL.phys.data, lakename, sampledate:temperature_C) #we want the sample date (year information), depth, and temperature). We don't have to list all of them because we can do : which allows us to select the range.

```
### Mutate

Mutating allows us to add new columns that are functions of existing columns. Operations include addition, subtraction, multiplication, division, log, and other functions.

```{r}

NTL.phys.data.temps <- mutate(NTL.phys.data.temps, temperature_F = (temperature_C*9/5) + 32)
#imagine we want a column with temperature but on another unit. Mutate, specify original data set and the function of the rows. Temperature in F will be C * 9/5 plus 32. Generating new column that will have temperature in F. Mutating within the temps frame.

```

## Lubridate

A package that makes coercing date much easier is `lubridate`. A guide to the package can be found at https://lubridate.tidyverse.org/. The cheat sheet within that web page is excellent too. This package can do many things (hint: look into this package if you are having unique date-type issues), but today we will be using two of its functions for our NTL dataset. 

```{r}
# add a month column to the dataset
NTL.phys.data.PeterPaul1 <- mutate(NTL.phys.data.PeterPaul1, month = month(sampledate)) #helpful to aggregate data by month

# reorder columns to put month with the rest of the date variables
NTL.phys.data.PeterPaul1 <- select(NTL.phys.data.PeterPaul1, lakeid:daynum, month, sampledate:comments)
#select is allowing us to rearrange

# find out the start and end dates of the dataset
interval(NTL.phys.data.PeterPaul1$sampledate[1], NTL.phys.data.PeterPaul1$sampledate[21613]) #allows you to see first and last dates
interval(first(NTL.phys.data.PeterPaul1$sampledate), last(NTL.phys.data.PeterPaul1$sampledate))
#creates an interval object with specified start and end. If you want to select an interval within your data set...
#function first and last are from dplyr.if not ordered by date. in this finding first and last date.
```


## Pipes

Sometimes we will want to perform multiple functions on a single dataset on our way to creating a processed dataset. We could do this in a series of subsequent functions or create a custom function. However, there is another method to do this that looks cleaner and is easier to read. This method is called a pipe. We designate a pipe with `%>%`. A good way to think about the function of a pipe is with the word "then." 

Let's say we want to take our raw dataset (NTL.phys.data), *then* filter the data for Peter and Paul lakes, *then* select temperature and observation information, and *then* add a column for temperature in Fahrenheit: 

```{r}
NTL.phys.data.processed <- 
  NTL.phys.data %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake") %>%
  select(lakename, sampledate:temperature_C) %>%
  mutate(temperature_F = (temperature_C*9/5) + 32)
  
```

Notice that we did not place the dataset name inside the wrangling function but rather at the beginning.

### Saving processed datasets

```{r}
write.csv(NTL.phys.data.PeterPaul1, row.names = FALSE, file = "./Data/Processed/NTL-LTER_Lake_ChemistryPhysics_PeterPaul_Processed.csv")
```

## Closing Discussion

When we wrangle a raw dataset into a processed dataset, we create a code file that contains only the wrangling code. We then save the processed dataset as a new spreadsheet and then create a separate code file to analyze and visualize the dataset. Why do we keep the wrangling code separate from the analysis code?

At the end of wrangling, we want to save. We could just do visualization next, or answer question. Or we could combine. Processing of first data set is done, close. Start a new one because we don't want to do everything in same document. Then you will have really long RMD file, you will get lost. Good practice to wrangle into processed, create the code for wrangling, save as new spreadsheet, and then create a separate code to analyze and visualize. Simple as possible for reproducability and sharing.

