---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Hanna Bliska"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=45), tidy=TRUE)
getwd()
library(tidyverse)
library(agricolae)
library(lubridate)

#importing raw data with read.csv
NTL.LTER.Lake.Chem.Phys <-read.csv(
  "./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", 
  stringsAsFactors = TRUE)

#using as.date to format my dates
NTL.LTER.Lake.Chem.Phys$sampledate <- as.Date(
  NTL.LTER.Lake.Chem.Phys$sampledate , format = "%m/%d/%y")

#2
#building my theme
mytheme <- theme_classic(base_size = 11) + theme(
  axis.text = element_text(color="black"), 
  legend.position = "right")
theme_set(mytheme) #setting my theme

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The slope of the linear regression between depth and mean lake temperature is equal to zero; thus, depth has no effect on mean lake temperature.
Ha: The slope of the linear regression between depth and mean lake temperature is not equal to zero; thus, depth has an effect on mean lake temperature.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
Processed.NTL.LTER <- 
  NTL.LTER.Lake.Chem.Phys %>%
  mutate(month = month(sampledate)) %>% #creating a month column
  filter(month=="7") %>% #including only July
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  #selecting only these columns
  drop_na(lakename, year4, daynum, depth, temperature_C) #using
  #drop_na to remove NAs in my columns
  
#5
Plot.Depth.Temp <- ggplot(
  Processed.NTL.LTER, aes(x=temperature_C, y=depth)) +
  geom_point() + #creating a scatter plot
  ylab(expression("Temperature (C)")) + #setting y axis label
  xlab(expression("Depth (m)")) + #setting x axis label
  ylim(0,35) + #setting limit for y axis
  geom_smooth(method=lm, se = FALSE, color="red") #adding linear model
print(Plot.Depth.Temp)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This figure suggests that temperature decreases in response to increased depths. This is an intuitive result, as the load of light energy (which heats water) is high at shallow depths but decreases as it travels through water to deeper depths. The distribution of points suggests that the relationship between depth and temperature is a decreasing linear trend, however, there is some variation in the distribution. The variation suggests that the strength of the linear relationship may not be as strong, which we can test through performing a linear regression.

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
temp.depth.regression <- lm(
  data = Processed.NTL.LTER, temperature_C ~ depth) 
#linear regression between temperature and depth
summary(temp.depth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: 73.87% of the variability in temperature is explained by changes in depth, as indicated by the R^2 of my model (0.7387). The degrees of freedom for my model are 9726, which is the number of observations minus two (for two variables). The relationship between depth and temperature is negative and statistically significant, as the p value of my model is less than 0.05. This indicates that deeper depths have significantly lower temperatures than shallower depths. Temperature is predicted to decrease -1.94621 degrees celcius for every 1m deeper change in depth.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
AIC.Processed.NTL.LTER <- 
  lm(data = Processed.NTL.LTER, temperature_C ~ depth + daynum + 
              year4) #considering all explanatory variables in AIC
step(AIC.Processed.NTL.LTER)

#10
temp.mult.regression <-
  lm(data = Processed.NTL.LTER, temperature_C ~ depth + daynum + 
              year4) 
#running a multiple regression with variables
#recommended by AIC
summary(temp.mult.regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables are depth, daynum, and year4. This multiple regression model explains 74.12% of the variance in temperature, as indicated by the R^2 of 0.7412). This is a slight improvement in variance over using only depth, which explained 73.87% of the variance in temperature. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#running anova between temperature of lakes and lake name with aov
Lake.temp.anova <- aov(
  data = Processed.NTL.LTER, temperature_C ~ lakename)
summary(Lake.temp.anova)

#running anova between temperature of lakes and lake name with lm
Lake.temp.anova2 <- lm(
  data = Processed.NTL.LTER, temperature_C ~ lakename)
summary(Lake.temp.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the lakes (p<0.001).

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
Plot.Temp.Lakes <-
  ggplot(Processed.NTL.LTER, aes(x=depth, y=temperature_C)) +
  geom_point(aes(color=lakename),alpha=0.5) + #adding lake name
  #aesthetics and transparency to scatter plot
  geom_smooth(method=lm, se = FALSE, color="red") + #adding linear
  #model to the graph in red
  xlab(expression("Depth (m)")) + #setting x axis label
  ylab(expression("Temperature (C)")) + #setting y axis label
  ylim(0,35) + #setting limit for y axis
  labs(color="Lake name") #setting legend label
print(Plot.Temp.Lakes)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Lake.temp.anova) #creating Tukey HSD test

Lake.temp.anova.groups <- HSD.test(
  Lake.temp.anova, 
  "lakename", group=TRUE) 
#evaluating the groups from Tukey HSD test results
Lake.temp.anova.groups #printing the groups
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Ward Lake, Paul Lake, and Peter lake have the same mean temperature; their mean temperatures are not statistically different. No lake has a mean temperature that is statistically distinct from all other lakes; we see in the results of the test, that there is at least two lakes in each group (for a, n=2; for b, n=2; for c, n=3; for d, n=3; for e, n=3). 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could use a two sample t-test to explore if Peter and Paul Lake have distinct mean temperatures. The null hypothesis of this test would be that there is no difference between the mean of the lakes' temperatures. The alternative hypothesis would be that there is a significant difference betwen the temperatures of Peter and Paul Lake. 

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
Crampton.Ward.Lakes <- Processed.NTL.LTER %>%
  filter(lakename=="Crampton Lake" | lakename=="Ward Lake")

t.test(data=Crampton.Ward.Lakes, temperature_C ~ lakename)
```

>Answer: There is not a significant difference in lake temperatures between Crampton and Ward Lakes (p=0.2649). Therefore, we can infer that the mean lake temperatures of Crampton and Ward lakes in July are statiscally the same. This does match my answer for part 16, as Crampton and Ward lake were a part of the same group (b) produced by the Tukey HSD test and comparison of means.
