---
title: "Assignment 09: Data Scraping"
author: "Hanna Bliska"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
getwd()
library(tidyverse)
library(lubridate)
library(rvest)
library(scales)

mytheme <- theme_classic(base_size = 12) + theme(
  axis.text = element_text(color="black"), 
  legend.position = "right") #creating a theme

theme_set(mytheme) #setting my theme
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2

webpage <- read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021") #fetching contents into webpage object

webpage #viewing object

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3
water.system.name <- webpage %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text()
water.system.name

pswid <- webpage %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()
pswid

ownership <- webpage %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()
ownership
  
max.withdrawals.mgd <- webpage %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text()
max.withdrawals.mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4
max_withdrawals_df <- data.frame(
  "Month"=as.factor(c("Jan", "May", "Sept", "Feb", "Jun","Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")),
  "Year"=as.factor(rep(2021,12)),
  "Water System Name"=water.system.name,
  "PSWID"=pswid,
  "Ownership"=ownership,
  "Maximum Daily Withdrawals"=as.numeric(max.withdrawals.mgd))

max_withdrawals_df <- max_withdrawals_df %>%
  mutate(Date = my(paste(Month,"-",Year))) %>%
  arrange(ymd(Date))
 
#5
max_withdrawals_plot <- 
  ggplot(max_withdrawals_df, aes(
    x=Date, y=Maximum.Daily.Withdrawals))+ 
  geom_line() + #creating line plot
  scale_x_date(date_breaks="1 month", labels=date_format("%m-%Y")) +
  #using scale_x_date to make a break for each month in x axis
  theme(axis.text.x=element_text(angle=60, hjust=1)) +
  #tilting the angle of the x axis text
  ylab(expression("Maximum Daily Withdrawals (mgd)")) + 
  #setting y axis label
  xlab(expression("Date")) #setting x axis label
max_withdrawals_plot
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.
scrape.it <- function(the_PWSID, the_year){
  #retrieve website contents 
  the_website <- read_html(paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?",the_pwsid,"&",the_year))
  
  #set element address variables
  
  
  
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
#testing the function


#plotting

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8


```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019. Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9


```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?