---
title: "Assignment 2: Coding Basics"
author: "Hanna Bliska"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r Creating a sequence, echo=TRUE}
#1. 
seq_by_four <- seq(1, 100, 4) #from one, to one hundred, by fours

#2. 
mean(seq_by_four) #calculate mean of seq_by_four
median(seq_by_four) #calculate median of seq_by_four

#3. 
mean(seq_by_four) > median(seq_by_four) #asking R if the mean is greater than the median, output will return TRUE or FALSE

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r Creating a data frame, echo=TRUE}
student_names <- c("Hanna", "Caroline", "Isaac", "Sam") #this vector consists of characters
test_scores <- c(48, 90, 38, 95) #this vector consists of numbers
passing_scores <- c(FALSE, TRUE, FALSE, TRUE) #this vector consists of logical elements

df_student_names <- as.data.frame(student_names) #start by turning one vector into a dataframe
df_student_scores <- cbind(df_student_names, test_scores, passing_scores) #add columns to the dataframe to create a single data frame with all vectors

colnames(df_student_scores) <- c("student.first.name", "test.scores.numeric", "test.scores.passing") #renaming columns more informative names

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: This data frame includes elements from different classes (characters, numbers, logical elements). A matrix would include elements from the same class (e.g., only numbers). 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r Creating a function, echo=TRUE}
check_students_pass <- function(test_scores) {
  ifelse(test_scores>=50,TRUE,FALSE) #log_exp, if TRUE, if false
}
#this allowed me to create a function called check_students_pass which will use my test scores and return TRUE or FALSE depending on if the test score was less than 50

did_my_students_pass <-check_students_pass(test_scores) #using function check_students_pass on my test_scores
did_my_students_pass #output

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: When I tried to use the `if` and `else` option, I received an error that stated the condition has a length greater than 1. This is because I was asking `if` and `else` to evaluate all of the test scores in my vector, not a single test score. When I used `ifselse` I was able to evaluate all of the test scores in my vector. This is because `ifelse` can evaluate vectors.
