---
title: "Assignment 7: Time Series Analysis"
author: "Hanna Bliska"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on time series analysis. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A07_TimeSeries.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 16 at 11:59 pm.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#1
getwd() #checking wd

#loading packages
library(tidyverse)
library(lubridate)
library(zoo)
library(Kendall)
library(trend)
library(scales)

#setting my theme
mytheme <- theme_classic(base_size = 12) + theme(
  axis.text = element_text(color="black"), 
  legend.position = "right") #building my theme
theme_set(mytheme) #setting my theme

#2
#importing datasets
O3_2010 <- read.csv(
  "./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2010_raw.csv", stringsAsFactors = TRUE)

O3_2011 <- read.csv(
  "./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2011_raw.csv", stringsAsFactors = TRUE)

O3_2012 <- read.csv(
  "./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2012_raw.csv", stringsAsFactors = TRUE)

O3_2013 <- read.csv(
  "./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2013_raw.csv", stringsAsFactors = TRUE)

O3_2014 <- read.csv(
  "./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2014_raw.csv", stringsAsFactors = TRUE)

O3_2015 <- read.csv(
  "./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2015_raw.csv", stringsAsFactors = TRUE)

O3_2016 <- read.csv(
  "./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2016_raw.csv", stringsAsFactors = TRUE)

O3_2017 <- read.csv(
  "./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2017_raw.csv", stringsAsFactors = TRUE)

O3_2018 <- read.csv(
  "./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2018_raw.csv", stringsAsFactors = TRUE)

O3_2019 <- read.csv(
  "./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2019_raw.csv", stringsAsFactors = TRUE)

#using `rbind` to combine the datasets into single dataframe
GaringerOzone <- rbind(
  O3_2010, O3_2011, O3_2012, O3_2013, 
  O3_2014, O3_2015, O3_2016, O3_2017, 
  O3_2018, O3_2019)
```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3
#formatting date column to be a date class
GaringerOzone$Date <-as.Date(
  GaringerOzone$Date, format="%m/%d/%Y")

# 4
#using select to wrangle dataset for specific columns
GaringerOzone <- select(GaringerOzone, Date, Daily.Max.8.hour.Ozone.Concentration, DAILY_AQI_VALUE)

# 5
#using `as.data.frame` to create a data frame
#data frame has one column with every day from 2010-01-01 to
#2019-12-31
Days <- as.data.frame(seq(as.Date("2010-01-01"), as.Date("2019-12-31"),"days")) 

#renaming column
colnames(Days) <- c("Date")

# 6
#using `left_join` to join the datasets
#listing Days first to preserve the number of rows
#by date because this is the column in common
GaringerOzone <- left_join(Days, GaringerOzone, by = c("Date"))

```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7
Ozone_Line_Plot <- ggplot(GaringerOzone, aes(x=Date, y=Daily.Max.8.hour.Ozone.Concentration)) +
  geom_line() +
  scale_x_date(date_breaks="1 year", labels=date_format("%Y")) +
  xlab(expression("Date")) +
  ylab(expression ("Maximum Ozone Concentration (ppm)")) +
  geom_smooth(method=lm, se = FALSE, color="red")
print(Ozone_Line_Plot)
```

>Answer: The linear trend of my data suggests that ozone concentrations may have slightly decreased over time, but that there is no significant upward or downward trend. However, the distribution of my data suggests that there may be seasonal variation in my data, as shown by the up and down movement in the data that appears periodically.

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8
GaringerOzone$Daily.Max.8.hour.Ozone.Concentration <- 
  na.approx(
  GaringerOzone$Daily.Max.8.hour.Ozone.Concentration)

```

> Answer: We used a linear interpolation because we wanted to fill our NA values with ozone concentrations that fall between the previous and next measurement of ozone concentrations. This makes sense given that we have relatively few, and infrequent, NAs. We did not use the piecewise constant method because this method would replace NAs with values equal to the measurement made nearest to that date, which would cause replicates in our ozone concentrations between two days rather than our desired measurement that would fall between the concentrations of two days and follows the trend of our data. Similarly, we did not use the spline interpolation because it uses a quadratic function to interpolate NAs; given the up and down linear distribution of our data, it is more logical to use a linear function to interpolate NAs rather than a quadratic function. 


9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r}
#9
#new dataframe with aggregate data
GaringerOzone.monthly <-
  GaringerOzone %>%
  mutate(month = month(Date)) %>% #adding month column
  mutate(year = year(Date)) %>% #adding a year column
  group_by(month, year) %>% #grouping by month and year 
  summarise(meanOzoneConcentration = mean(Daily.Max.8.hour.Ozone.Concentration)) #generating means

#separate line of code for month-year combinations
#pipe to create a new column combining month and year columns
GaringerOzone.monthly <-
  GaringerOzone.monthly %>%
  mutate("firstmonth"=my(paste0(month,"-",year)))

```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10
Daily.GaringerOzone.ts <- ts(
  GaringerOzone$Daily.Max.8.hour.Ozone.Concentration,
  start=c(2010,1,1),end=c(2019,12,31),frequency=365) 
#daily measurements, using 365

Monthly.GaringerOzone.ts <- ts(
  GaringerOzone.monthly$meanOzoneConcentration, 
  start=c(2010,1,1),end=c(2019,12,1),frequency=12) 
#monthly measurements, using 12

```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11
Daily.GaringerOzone.ts.Decomposed <- stl(
  Daily.GaringerOzone.ts, s.window = "periodic") 
#decomposing daily time series
plot(Daily.GaringerOzone.ts.Decomposed) #plotting components

Monthly.GaringerOzone.ts.Decomposed <- stl(
  Monthly.GaringerOzone.ts, s.window = "periodic")
#decomposing monthly time series
plot(Monthly.GaringerOzone.ts.Decomposed) #plotting components

```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12
Trend.Monthly.GaringerOzone <- 
  SeasonalMannKendall(
  Monthly.GaringerOzone.ts)
summary(Trend.Monthly.GaringerOzone)
```

> Answer: We ran a sesonal Mann-Kendall test because monthly ozone concentrations have  a seasonal cycle as demonstrated by the plot of the seasonal component produced in #11. In the plot, the largest bar (on the right hand side) is found in the the seasonal component, indicating that the seasonal component explains most of the variability in the data. A seasonal Mann-Kendall test is most appropriate for determining monotonic trends in data with seasonality.

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r}
# 13
Mean.Monthly.Ozone.Time.Plot <- ggplot(
  GaringerOzone.monthly, aes(
    x=firstmonth, y=meanOzoneConcentration)) +
  geom_point() +
  geom_line() +
  scale_x_date(date_breaks="6 months", labels=date_format("%d-%m-%Y")) +
  theme(axis.text.x=element_text(angle=60, hjust=1)) +
  xlab(expression("Date")) +
  ylab(expression ("Mean Ozone Concentration (ppm)"))
print(Mean.Monthly.Ozone.Time.Plot)

```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: The results of my trend analysis for monthly Ozone concentrations indicates that the seasonal trend is stationary (p=0.16323). This means that the trend is not decreasing or increasing. We can thus infer that the average monthly Ozone concentrations observed in the summer months (seasonally high periods of Ozone concentrations) have not risen over time compared to other summer months. Likewise, we can infer that the average monthly Ozone concentrations observed in the winter months (seasonally low periods of Ozone concentrations) have not decreased over time compared to other winter months.

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15
Monthly.GaringerOzone.NonseasonalComponents <- as.data.frame(Monthly.GaringerOzone.ts.Decomposed$time.series[,2:3])

Monthly.GaringerOzone.NonseasonalComponents <-
  mutate(Monthly.GaringerOzone.NonseasonalComponents, Observed=GaringerOzone.monthly$meanOzoneConcentration, Date=GaringerOzone.monthly$firstmonth)

Nonseasonal.ts <- ts(
  Monthly.GaringerOzone.NonseasonalComponents, 
  start=c(2010,1,1),end=c(2019,12,1),frequency=12) 

#16
Nonseasonal.Trend.Monthly.GaringerOzone <- 
  MannKendall(Nonseasonal.ts)
summary(Nonseasonal.Trend.Monthly.GaringerOzone)

```

> Answer: 
