---
title: "Assignment 5: Data Visualization"
author: "Hanna Bliska"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
#Hanna to check color, off of 0, units, how to make the legend smaller, month order, drop nas?
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 21st @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_
Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_
mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
library(tidyverse)
library(lubridate)
library(cowplot)
Peter.Paul.Nutrients.Chem <- read.csv(
  "./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv",
  stringsAsFactors = TRUE)
Niwot.Litter <- read.csv(
  "./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", 
  stringsAsFactors = TRUE)

#2 
Peter.Paul.Nutrients.Chem$sampledate <- as.Date(
  Peter.Paul.Nutrients.Chem$sampledate, 
  format = "%Y-%m-%d") #formatting date
Niwot.Litter$collectDate <- as.Date(
  Niwot.Litter$collectDate, format = "%Y-%m-%d") #formatting date

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
mytheme <- theme_classic(base_size = 12) + theme(
  axis.text = element_text(color="black"), 
  legend.position = "right") 
#building my theme 
#using classic theme with black text, size 13 text, 
#and legend position to the right
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4
Plot_TotalP_Phosphate <- ggplot(Peter.Paul.Nutrients.Chem, aes(x = po4, y = tp_ug)) +
  geom_point(aes(color=lakename), size=1) + #creating scatter plot
  xlim(0, 50) + #adjusting axis
  geom_smooth(method=lm, se = FALSE, color="black") +
  ylab(expression("Total Phosphorus")) + #setting y axis label
  xlab(expression("Phosphate")) + #setting x axis label
  labs(color="Lake name") + #setting legend label
  mytheme #utilizing my theme
print(Plot_TotalP_Phosphate)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
#5
Plot_Temp <- ggplot(Peter.Paul.Nutrients.Chem, aes(
  x = factor(month, levels=c(1:12)), y = temperature_C)) +
  geom_boxplot(aes(color=lakename), size=1) + #creating scatter plot
  scale_x_discrete(labels=month.abb, drop=FALSE) +
  ylab(expression("Temperature (C)")) + #setting y axis label
  xlab(expression("Month")) + #setting x axis label
  labs(color="Lake name") + #setting legend label
  mytheme #utilizing my theme
print(Plot_Temp)

Plot_TP <- ggplot(Peter.Paul.Nutrients.Chem, aes(
  x = fct_inseq(as.factor(month)), y = tp_ug)) +
  geom_boxplot(aes(color=lakename), size=1) +
  scale_x_discrete(labels=month) +
  ylab(expression("Total Phosphorus")) +
  xlab(expression("Month")) +
  labs(color="Lake name") +
  mytheme
print(Plot_TP)

Plot_TN <- ggplot(Peter.Paul.Nutrients.Chem, aes(
  x = fct_inorder(month.abb[month]), y = tn_ug)) +
  geom_boxplot(aes(color=lakename), size=1) +
  ylab(expression("Total Nitrogen")) +
  xlab(expression("Month")) +
  labs(color="Lake name") +
  mytheme
print(Plot_TN)

Plot_Temp_TP_TN <-
  plot_grid(
    Plot_Temp + theme(legend.position="none") + xlab(NULL),
    Plot_TP  + theme(legend.position="none") + xlab(NULL), 
    Plot_TN  + theme(legend.position="bottom"), 
            ncol=1, #all plots in one column
            nrows=3,
            align = 'v', #aligning in both directions 
            #rel_widths = c(10,10,10), #column widths
            rel_heights = c(1,1,1.5)) + #row heights
  mytheme

Plot_Temp_TP_TN_Legend <- plot_grid(Plot_Temp_TP_TN) #adding legend
print(Plot_Temp_TP_TN_Legend)
```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: It appears that the temperature of the lakes increases over the summer months with a peak in July and August. It also appears that nitrogen and phosphorus increase over the summer with peaks in August and September; there also appears to be a great amount of variability in the nutrient samples in the summer months. Peter Lake appears to have slightly higher concentrations of nitrogen and phosphorus than Paul Lake. Paul Lake appears to have slightly higher lake temperatures from May to Septmeber than Peter Lake. 

6. [Niwot Ridge] Plot a subset of the litter data set by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6
Litter_Plot_Needles <- ggplot(subset(
  Niwot.Litter, functionalGroup == "Needles"), aes(
    x=collectDate, y=dryMass)) + #subsetting functional group by needles
  geom_point(aes(color=nlcdClass)) + #creating scatter plot
  ylab(expression("Dry Mass of Needle Litter")) + #setting y axis label
  xlab(expression("Date of Collection")) + #setting x axis label
  labs(color="NLCD Class") + #setting legend label
  mytheme #utilizing my theme
print(Litter_Plot_Needles)

#7
Litter_Plot_Needles_Faceted <- ggplot(subset(
  Niwot.Litter, functionalGroup == "Needles"), aes(
    x=collectDate, y=dryMass)) + #subsetting functional group by needles
  geom_point() + #creating a scatter plot
  ylab(expression("Dry Mass of Needle Litter")) + #setting y axis label
  xlab(expression("Date of Collection")) + #setting x axis label
  facet_grid(functionalGroup ~ nlcdClass) + #using facet grid to facet the plot 
  #by NLCD class
  mytheme #utilizing my theme
print(Litter_Plot_Needles_Faceted)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think that plot 7 is more effective for visualizing the data because it allows the viewer to see the differences in the distribution of the dry mass data for each NLCD class more clearly. On plot 7, we can easily see that grassland herbaceous and evergreen forest classes had observations (potentially outliers) that were much higher in dry mass than the shrub scrub class. We are also more easily able to see the concentration in observations of dry mass in the shrub scrub class in Plot 7; this helps us to understand the tighter distribution of dry mass observations for this class as well as to understand that most of the observations were close to zero.  