---
title: "Assignment 5: Data Visualization"
author: "Hanna Bliska"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=45), tidy=TRUE)
#Hanna to check color, units
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 21st @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
library(tidyverse)
library(lubridate)
library(cowplot)
Peter.Paul.Nutrients.Chem <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
Niwot.Litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 
Peter.Paul.Nutrients.Chem$sampledate <- as.Date(Peter.Paul.Nutrients.Chem$sampledate, format = "%Y-%m-%d") #formatting date
Niwot.Litter$collectDate <- as.Date(Niwot.Litter$collectDate, format = "%Y-%m-%d") #formatting date

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
mytheme <- theme_classic(base_size = 13) + theme(axis.text = element_text(color="black"), legend.position = "right") #building my theme from the classic theme with black text, size 13 text, and legend position to the right
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4
Plot_TotalP_Phosphate <- ggplot(Peter.Paul.Nutrients.Chem, aes(x = po4, y = tp_ug)) +
  geom_point(aes(color=lakename), size=1) +
  xlim(0, 50) +
  geom_smooth(method=lm, se = FALSE, color="black") +
  ylab(expression("Total Phosphorus")) +
  xlab(expression("Phosphate")) +
  labs(color="Lake name") +
  mytheme
print(Plot_TotalP_Phosphate)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5
Plot_Temp <- ggplot(Peter.Paul.Nutrients.Chem, aes(x = month.abb[month], y = temperature_C)) +
  geom_point(aes(color=lakename), size=1) +
  ylab(expression("Temperature (C)")) +
  xlab(expression("Month")) +
  labs(color="Lake name") +
  mytheme
print(Plot_Temp)

Plot_TP <- ggplot(Peter.Paul.Nutrients.Chem, aes(x = month.abb[month], y = tp_ug)) +
  geom_point(aes(color=lakename), size=1) +
  ylab(expression("Total Phosphorus")) +
  xlab(expression("Month")) +
  labs(color="Lake name") +
  mytheme
print(Plot_TP)

Plot_TN <- ggplot(Peter.Paul.Nutrients.Chem, aes(x = month.abb[month], y = tn_ug)) +
  geom_point(aes(color=lakename), size=1) +
  ylab(expression("Total Nitrogen")) +
  xlab(expression("Month")) +
  labs(color="Lake name") +
  mytheme
print(Plot_TN)

Plot_Temp_TP_TN <-
  plot_grid(Plot_Temp, Plot_TP, Plot_TN, nrow = 3, align = 'h', rel_heights = c(10, 10)) +
  mytheme
print(Plot_Temp_TP_TN)
```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6

#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: